Unsupervised embedding of trajectories captures the latent structure of mobility

### 2020-10-05 paper/current/main.tex
conclusion edits

Neural embedding approaches offer a novel, data-driven solution for learning an effective and robust representation of locations based on trajectory data, encoding the complex and multi-faceted nature of mobility.
We demonstrated that a functional distance derived from the embedding can be used with the gravity model of mobility to better predicts real-world mobility than does geographic distance.
Embedding distance outperformed geographic distance both in the case of U.S. flight itineraries and global scientific mobility domains of mobility with very different characteristics. 
Focusing on scientific mobility, we find that neural embedding's performance is driven by its ability to encode many aspects of scientific mobility into a single representation, including global and regional geography, shared languages, and prestige hierarchies, even without explicit information on these factors.

While we focus on data of U.S. flight itineraries and scientific mobility, this approach may be broadly applicable to other kinds of mobility data, such as general human migration, transit-network mobility, and more.
Moreover, this approach can be used to learn a functional distance between entities for which no physical analog exists, such as between occupational categories based on individuals' career trajectories.
In addition to providing a functional distance that supports modeling and predicting mobility patterns, the structure of the neural embedding space is amenable to a range of unique applications for studying mobility.
As we have shown, the embedding space allows the visualization of the complex structure of scientific mobility at high resolution across multiple scales, providing a large and detailed map of the landscape of global scientific mobility.
Embedding also allows us to quantify abstract notions in mobility such as academic prestige, and can also be generalized to other abstract axes.
Investigation of the structure of the embedding space, such as the vector norm, reveals unique and universal patterns based on the organization's size and their vector norm that could be leveraged in future studies of mobility.

In spite of its promise, our study has several limitations.
First, the current skip-gram word2vec model cannot encode directional information, meaning that embedding will be less effective at capturing mobility for which directionality is critical.
Second, the neural embedding approach is most useful in cases of mobility between discrete geographic units such as between countries, cities, and businesses; this approach is less useful in the case of mobility between locations represented using geographic coordinates, such as in the modeling of animal movements.
Neither of these methodological limitations is insurmountable, and future work can aim to incorporate directionality and identify meaningful representations of continuous mobility data.
Finally, the case of scientific mobility presents domain-specific limitations.
Reliance on bibliometric metadata means that we capture only long-term mobility such as migration, rather than the array or more frequent short-term mobility such as conference travel and temporary visits.
The kinds of mobility we do capture---migration, and co-affiliation---although qualitatively different, are treated identically by our model.
Also, our data might further suffer from survivorship bias: researchers at prestigious organizations tend to have more publications, leading to these organizations appearing more frequently in affiliation trajectories.

Mobility and migration are at the core of human nature and history~\autocite{shah2020migration}, driving societal phenomena as diverse as epidemics\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity} and science and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage, morgan2018prestige, rodrigues2016mobility}.
However, the paradigm of scientific migration may be changing.
Traditional hubs of migration have experienced many politically-motivated policy changes that affect scientific mobility, such as travel restrictions in the U.S. and U.K.~\autocite{chinchilla2018travelban}.
At the same time, other nations, such as China, are growing into major scientific powers and attractors of talent~\autocite{tollefson2018china, cao2020returning}.
Unprecedented health crises such as the COVID-19 pandemic threaten to bring drastic global changes to travel and migration by tightening borders and halting travel.
With the changing paradigm of global mobility, we need now, more than ever, new tools and approaches to capture and understand human mobility in order to inform sensible, effective, sustainable, and humane policies.

>>>>

Neural embedding approaches offer a novel, data-driven solution for learning an effective and robust representation of locations based on trajectory data, encoding the complex and multi-faceted nature of mobility.
We demonstrated that a functional distance derived from the embedding can be used with the gravity model of mobility to better predicts real-world mobility than does geographic distance.
Embedding distance outperformed geographic distance both in the case of U.S. flight itineraries and global scientific mobility, which represent two domains with disparate nature. 
Focusing on scientific mobility, we find that neural embedding's performance is driven by its ability to encode many aspects of scientific mobility into a single representation, including global and regional geography, shared languages, and prestige hierarchies, even without explicit information on these factors.

While we focus on data of U.S. flight itineraries and scientific mobility, this approach may be broadly applicable to other kinds of mobility data, such as general human migration, transit-network mobility, and more.
Moreover, this approach can be used to learn a functional distance even between entities for which no physical analog exists, such as between occupational categories based on individuals' career trajectories.
In addition to providing a functional distance that supports modeling and predicting mobility patterns, the structure of the neural embedding space is amenable to a range of unique applications for studying mobility.
As we have shown, the embedding space allows the visualization of the complex structure of scientific mobility at high resolution across multiple scales, providing a large and detailed map of the landscape of global scientific mobility.
Embedding also allows us to quantitatively explore abstract notions such as academic prestige, and can potentially be generalized to other abstract axes.
Investigation of the structure of the embedding space, such as the vector norm, reveals universal patterns based on the organization's size and their vector norm that could be leveraged in future studies of mobility.

In spite of its promise, our study has several limitations.
First, the skip-gram word2vec model does not leverage directionality, meaning that embedding will be less effective at capturing mobility for which directionality is critical.
Second, the neural embedding approach is most useful in cases of mobility between discrete geographic units such as between countries, cities, and businesses; this approach is less useful in the case of mobility between locations represented using geographic coordinates, such as in the modeling of animal movements.
Neither of these methodological limitations is insurmountable, and future work can aim to incorporate directionality and identify meaningful representations of continuous mobility data.
Finally, the case of scientific mobility presents domain-specific limitations.
Reliance on bibliometric metadata means that we capture only long-term mobility such as migration, rather than the array or more frequent short-term mobility such as conference travel and temporary visits.
The kinds of mobility we do capture---migration, and co-affiliation---although qualitatively different, are treated identically by our model.
Also, our data might further suffer from survivorship bias: researchers at prestigious organizations tend to have more publications, leading to these organizations appearing more frequently in affiliation trajectories.

Mobility and migration are at the core of human nature and history~\autocite{shah2020migration}, driving societal phenomena as diverse as epidemics\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity} and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage, morgan2018prestige, rodrigues2016mobility}.
However, the paradigm of scientific migration may be changing.
Traditional hubs of migration have experienced many politically-motivated policy changes that affect scientific mobility, such as travel restrictions in the U.S. and U.K.~\autocite{chinchilla2018travelban}.
At the same time, other nations, such as China, are growing into major scientific powers and attractors of talent~\autocite{tollefson2018china, cao2020returning}.
Unprecedented health crises such as the COVID-19 pandemic threaten to bring drastic global changes to travel and migration by tightening borders and halting travel.
With the changing paradigm of global mobility, we need now, more than ever, new tools and approaches to capture and understand human mobility in order to inform sensible, effective, sustainable, and humane policies.
